Refinement of Parallel Algorithms down to LLVM

### 2023-03-23 papers/JAR_SI_ITP2022/main.tex
JAR SI ITP paper

\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
%This is a template for producing LIPIcs articles.
%See lipics-v2021-authors-guidelines.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling autoref support, use "autoref"
%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
%for enabling thm-restate support, use "thm-restate"
%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
%for producing a PDF according the PDF/A standard, add "pdfa"

%\pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository

%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory

%
% \documentclass[sigplan,10pt,anonymous,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}

\bibliographystyle{plainurl}% the mandatory bibstyle

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{microtype} % Better typesetting for PDFs -- is enabling this ok?
\usepackage{balance}

\usepackage{amsmath}
\usepackage{amssymb}
% \usepackage{amsthm}

%\usepackage{eufrak} %The eufrak package is redundant if the amsfonts package is used
\usepackage{mathpartir}
\DeclareMathAlphabet{\mathpzc}{OT1}{pzc}{m}{it}
\usepackage[boxed]{algorithm}
\usepackage{enumerate}
\usepackage{cprotect}

\usepackage{listings}
\usepackage{lstautogobble}
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{ragged2e}  % for '\RaggedRight' macro (allows hyphenation)
\usepackage{color}
\usepackage[noend]{algpseudocode}
\usepackage{caption}
\usepackage[font=scriptsize]{subcaption}
\usepackage{hyperref}
\usepackage{float}
% \usepackage{wrapfig}

\usepackage{multirow}
\usepackage{siunitx}

@@

% \theoremstyle{definition}
% \newtheorem{example}{Example}[section]

% \overfullrule=8pt

\title{Refinement of Parallel Algorithms down to LLVM}

% \author{Peter Lammich}
% \institute{University of Twente}
% \author{A.~Nonymous}
% \institute{Institute}
% \email{p.lammich@utwente.nl}
%
% \authorinfo{Peter Lammich}
%            {The University of Manchester, UK}
%            {peter.lammich@machester.ac.uk}

\author{Peter Lammich}{University of Twente, Netherlands}{p.lammich@utwente.nl}{https://orcid.org/0000-0003-3576-0504}{}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty. Please provide at least the name of the affiliation and the country. The full address is optional. Use additional curly braces to indicate the correct name splitting when the last name consists of multiple name parts.

\authorrunning{Peter Lammich} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

\Copyright{Peter Lammich} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

\ccsdesc[500]{Software and its engineering~Formal software verification}
\ccsdesc[500]{Theory of computation~Semantics and reasoning}
\ccsdesc[300]{Computing methodologies~Parallel algorithms}

% \ccsdesc[100]{\textcolor{red}{Replace ccsdesc macro with valid one}} %TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm

\keywords{Isabelle,Concurrent Separation Logic,Parallel Sorting,LLVM} %TODO mandatory; please add comma-separated list of keywords

\supplementdetails[subcategory={Isabelle Formalization}]{Software}{https://www21.in.tum.de/~lammich/isabelle_llvm_par/}

% \acknowledgements{I want to thank \dots}%optional

%\nolinenumbers %uncomment to disable line numbering

% %Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \EventEditors{John Q. Open and Joan R. Access}
% \EventNoEds{2}
% \EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
% \EventShortTitle{CVIT 2016}
% \EventAcronym{CVIT}
% \EventYear{2016}
% \EventDate{December 24--27, 2016}
% \EventLocation{Little Whinging, United Kingdom}
% \EventLogo{}
% \SeriesVolume{42}
% \ArticleNo{23}
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\EventEditors{June Andronick and Leonardo de Moura}
\EventNoEds{2}
\EventLongTitle{13th International Conference on Interactive Theorem Proving (ITP 2022)}
\EventShortTitle{ITP 2022}
\EventAcronym{ITP}
\EventYear{2022}
\EventDate{August 7--10, 2022}
\EventLocation{Haifa, Israel}
\EventLogo{}
\SeriesVolume{237}
\ArticleNo{12}

@@

@@

\begin{abstract}
  We present a stepwise refinement approach to develop
  verified parallel algorithms, down to efficient LLVM code.
  The resulting algorithms' performance is competitive with their
  counterparts implemented in C/C++.
%
  Our approach is backwards compatible with the Isabelle Refinement Framework,
  such that existing sequential formalizations can easily be adapted
  or re-used.
%
  As case study, we verify a parallel quicksort algorithm, and
  show that it performs on par with its C++ implementation,
  and is competitive to state-of-the-art parallel sorting algorithms.
\end{abstract}

@@

We present a stepwise refinement approach to develop
verified and efficient parallel algorithms. Our method can verify total correctness down to LLVM intermediate code.
The resulting verified implementations are competitive with state-of-the-art
unverified implementations. Our approach is backwards compatible to
the Isabelle Refinement Framework (IRF), a powerful tool to verify efficient sequential software,
such as model checkers~\cite{ELNN13,BrLa18,WiLa18}, SAT solvers~\cite{La17_CADE,La17_SAT,FBL18},
or graph algorithms~\cite{La14,LaSe16,LaSe19}.
This paper adds parallel execution to the IRF's toolbox, without invalidating the existing formalizations,
which can now be used as sequential building blocks for parallel algorithms, or be modified to add
parallelization.

As a case study, we verify total correctness of a parallel quicksort algorithm,
re-using an existing verification of state-of-the-art sequential sorting algorithms~\cite{La20}.
Our verified parallel sorting algorithm is competitive to state-of-the-art parallel sorting algorithms.

@@

% \noindent The formalization is available at \url{https://www21.in.tum.de/~lammich/isabelle_llvm_par/}.

\subsection{Overview}\label{sec:intro_overview}
This paper is based on the Isabelle Refinement Framework, a continuing effort
to verify efficient implementations of complex algorithms, using stepwise refinement techniques.
Figure~\ref{fig:IRF-overview} displays the components of the Isabelle Refinement Framework.

\begin{figure}
  \definecolor{colA}{RGB}{11, 97,164}
  \definecolor{colB}{RGB}{204,246,  0}
  \definecolor{colC}{RGB}{166,  0,166}
  \definecolor{colD}{RGB}{255,146,  0}
  \definecolor{colTCB}{RGB}{255, 0, 0}

@@

  \caption{Components of the Isabelle Refinement Framework, with focus on the back end.}\label{fig:IRF-overview}
\end{figure}

%   \begin{tikzpicture}[node distance=.01\textwidth,outer sep=0pt,
%   layerA/.style={
%     draw=black,
%     fill=blue!20,
%     rounded corners,
%     text width=.9\textwidth,
%     font={\sffamily\bfseries\color{black}},
%     align=center,
%     minimum height=.03\textheight,
% %    text height=12pt,
% %    text depth=9pt
%     },
%   layerB/.style={layerA,fill=green!20},
%   layerC/.style={layerA,fill=red!20},
%   layerD/.style={layerA,fill=red!30,text width=.2\textwidth},
%   ]
%
%   \node[layerA] (AlgDS) {Algorithms and Data Structures};
%
%   \node[layerB, below=of AlgDS.south west, anchor=north west, text width=.437\textwidth] (ne) {Nres-Monad};
%   \node[layerA, right=of ne.north east, anchor=north west, text width=.437\textwidth, minimum height=.067\textheight] (llDS) {Low-Level Algorithms \\and Data Structures};
%
%   \node[layerB, below=of ne.south west, anchor=north west, text width=.437\textwidth] (sepref) {Sepref};
%
%   \node[layerC, below=of sepref.south west, anchor=north west] (sepl) {Program Logic and VCG};
%   \node[layerC, below=of sepl.south west, anchor=north west, minimum height=.120\textheight] (sembox) {};
%
%   \node[below right=of sembox.north west] {Back End Semantics};
%
%
%   \node[layerD, right=of sembox.west] (isacode) {Isa Codegen};
%   \node[layerD, right=of isacode] (imphol) {Imp HOL};
%   \node[layerD, right=of imphol] (llvm) {LLVM};
%   \node[layerD, right=of llvm] (llvm-par) {LLVM par};
%
%   \node[layerD, below=of imphol] (impholtime) {Imp HOL time};
%   \node[layerD, below=of llvm] (llvmtime) {LLVM time};
%
%
%   \end{tikzpicture}

The back end layer handles the translation from Isabelle/HOL to the actual target language.
The instructions of the target language are shallowly embedded into Isabelle/HOL, using
a state-error (SE) monad. An instruction with undefined behaviour,
or behaviour outside our supported fragment, raises an error.
The state of the monad is the memory, represented via a memory model.
The code generator translates the instructions to actual code.
These components form the trusted code base, while all the remaining components of the
Isabelle Refinement Framework generate proofs. In the back-end, the preprocessor transforms
expressions to the syntactically restricted format required by the code generator,
proving semantic equality of the original and transformed expression.
While there exist back ends for purely functional code~\cite{LaTu12,La13},
and sequential imperative code~\cite{La15,La19-llvm}, this paper describes a back end
for parallel imperative LLVM code (Section~\ref{sec:parsem}).

@@

At the level of the program logic and VCG, our framework can already be used to verify simple low-level algorithms and data structures,
like dynamic arrays and linked lists. More complex developments typically use a stepwise refinement approach,
starting at purely functional programs modelled in a nondeterminism-error (NE) monad~\cite{LaTu12}.
A semi-automatic refinement procedure (Sepref~\cite{La15,La19-llvm}) translates from the purely functional
code to imperative code, refining abstract functional data types to concrete imperative ones.
In Section~\ref{sec:refine_parallel}, we describe our extensions to support refinement to parallel executions,
and a fine-grained tracking of pointer equalities, required to parallelize computations that work on
disjoint parts of the same array.

@@

\section{A Back End for LLVM with Parallel Execution}\label{sec:parsem}
  We formalize a semantics for parallel execution, shallowly embedded into Isabelle/HOL.
  As for the existing sequential back ends~\cite{La15,La19-llvm}, the shallow embedding is key
  to the flexibility and feasibility of the approach.
  The main idea is to make an execution report the memory that it accesses,
  and use this information to raise an error when joining executions that
  would have exhibited a data race. We use this to model an instruction that
  calls two functions in parallel, and waits until both have returned.

  \subsection{State-Nondeterminism-Error Monad with Access Reports}\label{sec:monad}
  We define the underlying monad in two steps. We start with a nondeterminism-error monad,
  and then lift it to a state monad and add access reports.
  Defining a nondeterminism-error monad is straightforward in Isabelle/HOL:
  \begin{lstlisting}
    'a neM == spec ("'a => bool") | fail
    return x == spec (%r. r=x)
    bind fail f == fail
    bind (spec P) f == if ?x. P x & f x = fail then fail
                      else spec (%r. ?x Q. P x & f x = spec Q & Q r)
  \end{lstlisting}
  A program either fails, or yields a possible set of results (\is{spec P}), described by its characteristic function \is{P}.
  The \is{return} operation yields exactly one result, and \is{bind} combines all possible results,
  failing if there is a possibility to fail.

  Now assume that we have a state (memory) type \is{'\mu}, and an access report type \is{'\rho}, which forms a monoid (\is{0,+}).
  With this, we define our state-nondeterminism-error monad with access reports, just called \is{M} for brevity:
  \begin{lstlisting}
    'x M == '\mu => ('x \x '\rho \x '\mu) neM
    return$_M$ x \mu == return$_{ne}$ (x,0,\mu)
    bind$_M$ m f \mu == (x_1,r_1,\mu) <- m \mu; (x_2,r_2,\mu) <- f x_1 \mu; return$_{ne}$ (x_2,r_1+r_2,\mu)
  \end{lstlisting}
  Here, \is{return} does not change the state, and reports no accesses (\is$0$),
  and \is{bind} sequentially composes the executions, threading through the state $\mu$, and adding up the access reports \is$r_1$ and \is$r_2$.

  Typically, the access report will contain read and written addresses, such that data races can be detected.
  Moreover, if parallel executions can allocate memory, we must detect those
  executions where the memory manager allocated the same block in both parallel strands.
  As we assume a thread safe memory manager, those \emph{infeasible} executions can safely be ignored.
  Let \is{norace :: '\rho => '\rho => bool} and \is{feasible :: '\rho => '\rho => bool} be symmetric predicates,
  and let \is{combine :: ('\rho \x '\mu) => ('\rho \x '\mu) => ('\rho \x '\mu) } be a commutative operator to
  compose two pairs of access reports and states. Then, we define a parallel composition operator for \is{M}:
  \begin{lstlisting}
    (m_1 || m_2) \mu ==
      (x_1,r_1,\mu_1) <- m_1 \mu; (x_2,r_2,\mu_2) <- m_2 \mu;  --"\rightcomment{execute both strands}"
      assume feasible \rho_1 \rho_2;  --"\rightcomment{ignore infeasible combinations}"
      assert norace \rho_1 \rho_2; --"\rightcomment{fail on data race}"
      return$_{ne}$ ((x_1,x_2), combine (\rho_1,\mu_1) (\rho_2,\mu_2))  --"\rightcomment{combine results}"

    assume P == if P then return () else spec (%_. False)
    assert P == if P then return () else fail
  \end{lstlisting}
  Here, we use \is{assume} to ignore infeasible executions, and \is{assert} to fail on data races.
  Note that, if one parallel strand fails, and the other parallel strand has no possible
  results \is{spec (\<lambda>_. False)}, the behaviour of the parallel composition is not clear.
  For this reason, we fix an invariant \is{invar$_M$ :: ('\mu => ('x \x '\rho \x '\mu) neM) => bool},
  which implies that every non-failing execution has at least one possible result.
  We define the actual type \is{M} as the subtype satisfying \is{invar$_M$}.
  Thus, we have to prove that every combinator and instruction of our semantics
  preserves the invariant, which is an important sanity check.
  As additional sanity check, we prove symmetry of parallel composition:
  \begin{lstlisting}
    m_1 || m_2 = mswap (m_2 || m_1)     $\text{where}$    mswap m == (x_1,x_2)<-m; return (x_2,x_1)
  \end{lstlisting}
%   where \is|mswap m = (x_1,x_2)<-m; return (x_2,x_1)| swaps the result tuple.

@@

  \subsection{Access Reports}
  We now fix the state of the M-monad to be memory, and the access reports to be sets
  of read and written addresses, as well as sets of allocated and freed blocks:
  \begin{lstlisting}
    acc == ( r :: addr set; w :: addr set; a :: nat set; f :: nat set )
    0 == ( {},{},{},{} )
    (r_1,w_1,a_1,f_1) + (r_2,w_2,a_2,f_2) == ( r_1\union r_2, w_1\union w_2, a_1\union a_2, f_1\union f_2 )
  \end{lstlisting}

  Two parallel executions are feasible if they did not allocate the same block, and they have a data race
  if one strand accesses addresses or blocks modified by the other strand:
  \begin{lstlisting}
    feasible (r_1,w_1,a_1,f_1) (r_2,w_2,a_2,f_2) == a_1 \inter a_2 = {}

    norace (r_1,w_1,a_1,f_1) (r_2,w_2,a_2,f_2) ==
      let m_1 = w_1 \<union> { ADDR b i. b \<in> a_1 \<union> f_1 } in
      let m_2 = w_2 \<union> { ADDR b i. b \<in> a_2 \<union> f_2 } in
        (r_1 \<union> m_1) \<inter> m_2 = {}   &   m_1 \<inter> (r_2 \<union> m_2) = {}
  \end{lstlisting}
  The invariant for \is{M} states that blocks transition only from fresh to allocated to free,
  allocated blocks never change their size, and the access report matches the observable state change (\is{consistent}).
  It also states, that for each finite set of blocks $B$, there is an execution that does not allocate blocks from $B$.
  The latter is required to show that we always find feasible parallel executions:
  \begin{lstlisting}
    invar$_M$ c == \<forall>\mu P. c \mu = spec P ==>
      (\<forall>x \rho \mu'. P (x,\rho,\mu') ==> consistent \mu \rho \mu')
    & (\<forall>B. finite B ==> (\<exists>x \rho \mu'. P (x,\rho,\mu') & \rho.a \inter B = {} ))
  \end{lstlisting}
  The combine function joins the access reports and memories,
  preferring allocated over fresh, and freed over allocated memory. When joining two allocated blocks, the written addresses
  from the access report are used to join the blocks. We skip the rather technical definition of combine, and just state the
  relevant properties: Let \is{\rho_1=(r_1,w_1,a_1,f_1)} and \is{\rho_2=(r_2,w_2,a_2,f_2)} be feasible and race free access reports,
  and \is{\mu_1, \mu_2} be memories that have evolved from a common memory \is{\mu}, consistently with
  the access reports \is{\rho_1, \rho_2}. Let \is{(\rho',\mu') = combine (\rho_1,\mu_1) (\rho_2,\mu_2)}, and \is{addr} a valid address in \is{\mu'}. Then
  \begin{lstlisting}
    (1)  \mu' b = FRESH <--> \mu b = FRESH \and b ~: a_1 \union a_2
    (2)  is_alloc (\mu' b) <--> (is_alloc (\mu b) \or b \<in> a_1 \union a_2) \and b ~: f_1 \union f_2
    (3)  \mu' b = FREED <--> \mu b = FREED \or b \in f_1 \union f_2

>>>>

% TODO:
%   try to prevent page breaks for shorter listings
%   update website, prominently link to html-thys covered in this paper

% \documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
\documentclass[sn-mathphys,Numbered]{sn-jnl}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}

\usepackage{graphicx}%
\usepackage{multirow}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{amsthm}%
\usepackage{mathrsfs}%
\usepackage[title]{appendix}%
\usepackage{xcolor}%
\usepackage{textcomp}%
\usepackage{manyfoot}%
\usepackage{booktabs}%
\usepackage[boxed]{algorithm}%
\usepackage{algorithmicx}%
\usepackage[noend]{algpseudocode}
\usepackage{listings}%

% \usepackage[T1]{fontenc}
% \usepackage[utf8]{inputenc}
% \usepackage{microtype} % Better typesetting for PDFs -- is enabling this ok?
% \usepackage{balance}

%\usepackage{eufrak} %The eufrak package is redundant if the amsfonts package is used
% \usepackage{mathpartir}
% \DeclareMathAlphabet{\mathpzc}{OT1}{pzc}{m}{it}
% \usepackage[boxed]{algorithm}
% \usepackage{enumerate}
\usepackage{cprotect}

\usepackage{lstautogobble}
\usepackage{tabularx}
\usepackage{ragged2e}  % for '\RaggedRight' macro (allows hyphenation)
\usepackage{caption}
\usepackage[font=scriptsize]{subcaption}
% \usepackage{hyperref}
% \usepackage{float}
% \usepackage{wrapfig}

% \usepackage{multirow}
% \usepackage{siunitx}

@@

\theoremstyle{thmstyleone}%
\newtheorem{theorem}{Theorem}%  meant for continuous numbers
%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
\newtheorem{proposition}[theorem]{Proposition}%
%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.

\theoremstyle{definition}%
\newtheorem{example}{Example}%
\newtheorem{remark}{Remark}%

\theoremstyle{thmstylethree}%
\newtheorem{definition}{Definition}%

\raggedbottom

@@

\title[Refinement of Parallel Algorithms down to LLVM]{Refinement of Parallel Algorithms down to LLVM}
\subtitle{applied to practically efficient parallel sorting}

\author*[1]{\fnm{Peter} \sur{Lammich}}\email{p.lammich@utwente.nl}
\affil*[1]{\orgdiv{Faculty of Electrical Engineering, Mathematics and Computer Science} \orgname{University of Twente} \orgaddress{\city{Enschede} \country{Netherlands}}}

\abstract{
  We present a stepwise refinement approach to develop
  verified parallel algorithms, down to efficient LLVM code.
  The resulting algorithms' performance is competitive with their
  counterparts implemented in C/C++.
%
  Our approach is backwards compatible with the Isabelle Refinement Framework,
  such that existing sequential formalizations can easily be adapted
  or re-used.
%
  As case study, we verify a parallel quicksort algorithm that is
  competitive to unverified state-of-the-art algorithms.
}

\keywords{Isabelle,Concurrent Separation Logic,Parallel Sorting,LLVM}

@@

@@

We present a stepwise refinement approach to develop
verified and efficient parallel algorithms. Our method can verify total correctness down to LLVM intermediate code.
The resulting verified implementations are competitive with state-of-the-art
unverified implementations. Our approach is backwards compatible to
the Isabelle Refinement Framework (IRF)~\cite{La19-llvm}, a powerful tool to verify efficient sequential software,
such as model checkers~\cite{ELNN13,BrLa18,WiLa18}, SAT solvers~\cite{La17_CADE,La17_SAT,FBL18},
or graph algorithms~\cite{La14,LaSe16,LaSe19}.
This paper adds parallel execution to the IRF's toolbox, without invalidating the existing formalizations,
which can now be used as sequential building blocks for parallel algorithms, or be modified to add
parallelization.

As a case study, we verify total correctness of a parallel quicksort algorithm,
re-using an existing verification of state-of-the-art sequential sorting algorithms~\cite{La20}.
Our verified parallel sorting algorithm is competitive to state-of-the-art parallel sorting algorithms
from GNU's C++ standard library and the Boost C++ Libraries.

This paper is an extended version of our ITP 2023 paper~\cite{La22}.
The main new contribution is a verified parallel partitioning algorithm,
which significantly improves the efficiency and scalability of our sorting algorithm from~\cite{La22}.
To this end, we added a description of the interval list data structure (Section~\ref{sec:mod_ds})
and the \is{with_idxs} combinator (Section~\ref{sec:array_split}), which are required by the parallel partitioner.
The parallel partitioning algorithm itself is described in Section~\ref{sec:ppar}, and Section~\ref{sec:benchmark}
contains updated and more extensive benchmarks.

@@

\noindent The formalization is available at \href{https://www21.in.tum.de/~lammich/isabelle_llvm_par/}{www21.in.tum.de/$\sim$lammich/isabelle\_llvm\_par}.

\subsection{Overview}\label{sec:intro_overview}
This paper is based on the Isabelle Refinement Framework, a continuing effort
to verify efficient implementations of complex algorithms, using stepwise refinement techniques~\cite{LaLo10,LaTu12,La13,La15,La19-llvm,HaLa21}.
Figure~\ref{fig:IRF-overview} displays the components of the Isabelle Refinement Framework.
%
\begin{figure}
  \definecolor{colA}{RGB}{11, 97,164}
  \definecolor{colB}{RGB}{204,246,  0}
  \definecolor{colC}{RGB}{166,  0,166}
  \definecolor{colD}{RGB}{255,146,  0}
  \definecolor{colTCB}{RGB}{255, 0, 0}

@@

  \caption{Components of the IRF, with focus on the back end.}\label{fig:IRF-overview}
\end{figure}
%
The back end layer handles the translation from Isabelle/HOL to the actual target language.
The instructions of the target language are shallowly embedded into Isabelle/HOL, using
a state-error (SE) monad. An instruction with undefined behaviour,
or behaviour outside our supported fragment, raises an error.
The state of the monad is the memory, represented via a memory model.
The code generator translates the instructions to actual code.
These components form the trusted code base, while all the remaining components of the
Isabelle Refinement Framework generate proofs. In the back-end, the preprocessor transforms
expressions to the syntactically restricted format required by the code generator,
proving semantic equality of the original and transformed expression.
While there exist back ends for purely functional code~\cite{LaTu12,La13},
and sequential imperative code~\cite{La15,La19-llvm}, this paper describes a back end
for parallel imperative LLVM code (Section~\ref{sec:parsem}).

@@

At the level of the program logic and VCG, our framework can be used to verify simple low-level algorithms and data structures,
like dynamic arrays and linked lists. More complex developments typically use a stepwise refinement approach,
starting at purely functional programs modelled in a nondeterminism-error (NE) monad~\cite{LaTu12}.
A semi-automatic refinement procedure (Sepref~\cite{La15,La19-llvm}) translates from the purely functional
code to imperative code, refining abstract functional data types to concrete imperative ones.
In Section~\ref{sec:refine_parallel}, we describe our extensions to support refinement to parallel executions,
and a fine-grained tracking of pointer equalities, required to parallelize computations that work on
disjoint parts of the same array.

@@

\subsection{Notation}
We mainly use Isabelle/HOL notation, with some shortcuts and adaptations for presentation in a paper.
In this section, we give examples of the more unusual notations: for implication, we always write \is{==>},
e.g., \is{\<forall>x. x<4 ==> x<=3} (in Isabelle, one must use \is{-->} here).

Type variables are \is{'a, 'b, ...} and a function types are written curried as \is{'a => 'b => 'c}.
Types can be annotated to any symbol or term, e.g., \is{(x::nat) < 4 ==> x<=3}.
Function application is written as \is{f x}, and function update is \is{f(x:=y)}.
For definitions, we use \is{==}, e.g., \is{f x == x+1}.

Algebraic datatypes are written as \is{'a option == Some (the: 'a) | None}. This also defines the selector function \is{the :: 'a option => 'a}.
Values of the tuple type \is{'a_1 \x ... \x 'a_n} are written as \is{(a_1,...,a_n)}. Lists have type \is{'a list},
and we write \is{[]} for the empty list, \is{x#xs} for the list with head \is{x} and tail \is{xs}, and \is{xs_1@xs_2} for list concatenation.
We also use the list notation \is{[1,2,3]}. Then length of list \is{xs} is \is{|xs|}.

The empty set is \is${} :: 'a set$. The set \is${l..<h}$ contains all elements between $l$ inclusive and $h$ exclusive, and the set \is${l..}$ contains all elements greater than or equal to $l$.
Disjoint union is written as \is{$\dot\cup$}. We only write \is{$\dot\cup$} if it is clear from the context where the disjointness constraint belongs to.
Otherwise, we will explicitly write, e.g. \is$ s = s_1 \union s_2  &  s_1 \inter s_2 = {}$. The cardinality of the finite set \is{s} is \is{|s|}.

Lambda abstraction is written as \is{\<lambda>x. x+1}. When clear from the context, we omit the \is{\<lambda>}, e.g. \is{spec x. x>42}.
We also use Haskell-like sections for infix operators, e.g. \is{(+1)} for \is{\<lambda>x. x+1}.
Also, we can use underscores to indicate the parameter positions, e.g., \is${_..<_}$ for \is$\<lambda>l h. {l..<h}$.

\section{A Back End for LLVM with Parallel Execution}\label{sec:parsem}
  We formalize a semantics for parallel execution, shallowly embedded into Isabelle/HOL.
  As for the existing sequential back ends~\cite{La15,La19-llvm}, the shallow embedding is key
  to the flexibility and feasibility of the approach.
  The main idea is to make an execution report its accessed memory,
  and use this information to raise an error when joining executions that
  would have exhibited a data race. We use this to model an instruction that
  calls two functions in parallel, and waits until both have returned.

  \subsection{State-Nondeterminism-Error Monad with Access Reports}\label{sec:monad}
  We define the underlying monad in two steps. We start with a nondeterminism-error monad,
  and then lift it to a state monad and add access reports.
  Defining a nondeterminism-error monad is straightforward in Isabelle/HOL:
  \begin{lstlisting}
    'a neM == spec ("'a => bool") | fail
    return x == spec (%r. r=x)
    bind fail f == fail
    bind (spec P) f == if ?x. P x & f x = fail then fail
                            else spec (%r. ?x Q. P x & f x = spec Q & Q r)
  \end{lstlisting}
  A program either fails, or yields a possible set of results (\is{spec P}), described by its characteristic function \is{P}.
  The \is{return} operation yields exactly one result, and \is{bind} combines all possible results,
  failing if there is a possibility to fail. We use the notation \is{x<-m; f x} for \is{bind m (\<lambda>x. f x)},
  and \is{m_1; m_2} for \is{bind m_1 (\<lambda>_. m_2)}.

  Now assume that we have a state (memory) type \is{'\mu}, and an access report type \is{'\rho}, which forms a monoid (\is{0,+}).
  With this, we define our state-nondeterminism-error monad with access reports, just called \is{M} for brevity:
  \begin{lstlisting}
    'x M == '\mu => ('x \x '\rho \x '\mu) neM
    return$_M$ x \mu == return$_{ne}$ (x,0,\mu)
    bind$_M$ m f \mu == (x_1,r_1,\mu) <- m \mu; (x_2,r_2,\mu) <- f x_1 \mu; return$_{ne}$ (x_2,r_1+r_2,\mu)
  \end{lstlisting}
  Here, \is{return} does not change the state, and reports no accesses (\is$0$),
  and \is{bind} sequentially composes the executions, threading through the state $\mu$ and adding up the access reports \is$r_1$ and \is$r_2$.

  Typically, the access report will contain read and written addresses, such that data races can be detected.
  Moreover, if parallel executions can allocate memory, we must detect those
  executions where the memory manager allocated the same block in both parallel strands.
  As we assume a thread safe memory manager, those \emph{infeasible} executions can safely be ignored.
  Let \is{feasible :: '\rho => '\rho => bool} and \is{norace :: '\rho => '\rho => bool} be symmetric predicates,
  and let \is{combine :: ('\rho \x '\mu) => ('\rho \x '\mu) => ('\rho \x '\mu) } be a commutative operator to
  combine two pairs of access reports and states. Then, we define a parallel composition operator for \is{M}:
  \begin{lstlisting}
    (m_1 || m_2) \mu ==
      (x_1,r_1,\mu_1) <- m_1 \mu; (x_2,r_2,\mu_2) <- m_2 \mu;  --"\rightcomment{execute both strands}"
      assume feasible \rho_1 \rho_2;  --"\rightcomment{ignore infeasible combinations}"
      assert norace \rho_1 \rho_2; --"\rightcomment{fail on data race}"
      return$_{ne}$ ((x_1,x_2), combine (\rho_1,\mu_1) (\rho_2,\mu_2))  --"\rightcomment{combine results}"

    assume P == if P then return () else spec (%_. False)
    assert P == if P then return () else fail
  \end{lstlisting}
  Here, we use \is{assume} to ignore infeasible executions, and \is{assert} to fail on data races.
  Note that, if one parallel strand fails, and the other parallel strand has no possible
  results (\is{spec (\<lambda>_. False)}), the behaviour of the parallel composition is not clear.
  For this reason, we fix an invariant \is{invar$_M$ :: ('\mu => ('x \x '\rho \x '\mu) neM) => bool},
  which implies that every non-failing execution has at least one possible result.
  We define the actual type \is{M} as the subtype satisfying \is{invar$_M$}.
  Thus, we have to prove that every combinator and instruction of our semantics
  preserves the invariant, which is an important sanity check.
  As additional sanity check, we prove symmetry of parallel composition:
  \begin{lstlisting}
    m_1 || m_2 = mswap (m_2 || m_1)     $\text{where}$    mswap m == (x_1,x_2)<-m; return (x_2,x_1)
  \end{lstlisting}
%   where \is|mswap m = (x_1,x_2)<-m; return (x_2,x_1)| swaps the result tuple.

@@

  \subsection{Access Reports}
  We now fix the state of the M-monad to be memory, and the access reports to be tuples \is{(r,w,a,f)} of
  read and written addresses, as well as sets of allocated and freed blocks:
  \begin{lstlisting}
    acc == addr set \x addr set \x nat set \x nat set
    0 == ( {},{},{},{} )
    (r_1,w_1,a_1,f_1) + (r_2,w_2,a_2,f_2) == ( r_1\union r_2, w_1\union w_2, a_1\union a_2, f_1\union f_2 )
  \end{lstlisting}

  Two parallel executions are feasible if they did not allocate the same block. They have a data race
  if one execution accesses addresses or blocks modified by the other:
  \begin{lstlisting}
    feasible (r_1,w_1,a_1,f_1) (r_2,w_2,a_2,f_2) == a_1 \inter a_2 = {}

    norace (r_1,w_1,a_1,f_1) (r_2,w_2,a_2,f_2) ==
      let m_1 = w_1 \<union> { ADDR b i. b \<in> a_1 \<union> f_1 } in
      let m_2 = w_2 \<union> { ADDR b i. b \<in> a_2 \<union> f_2 } in
        (r_1 \<union> m_1) \<inter> m_2 = {}   &   m_1 \<inter> (r_2 \<union> m_2) = {}
  \end{lstlisting}
  The combine function joins the access reports and memories,
  preferring allocated over fresh, and freed over allocated memory. When joining two allocated blocks, the written addresses
  from the access report are used to join the blocks. We skip the rather technical definition of combine, and just state the
  relevant properties: Let \is{\rho_1=(r_1,w_1,a_1,f_1)} and \is{\rho_2=(r_2,w_2,a_2,f_2)} be feasible and race free access reports,
  and \is{\mu_1, \mu_2} be memories that have evolved from a common memory \is{\mu}, consistently with
  the access reports \is{\rho_1, \rho_2}. Let \is{(\rho',\mu') = combine (\rho_1,\mu_1) (\rho_2,\mu_2)}. Then
  \begin{lstlisting}
    (1)  \mu' b = FRESH <--> \mu b = FRESH \and b ~: a_1 \union a_2
    (2)  is_alloc (\mu' b) <--> (is_alloc (\mu b) \or b \<in> a_1 \union a_2) \and b ~: f_1 \union f_2
    (3)  \mu' b = FREED <--> \mu b = FREED \or b \in f_1 \union f_2
  \end{lstlisting}
  Moreover, for all addresses \is{addr = ADDR b i} with \is{valid \mu' addr}:
  \begin{lstlisting}
    (4)  addr \in w_1 \or b \in a_1 ==> get \mu' addr = get \mu_1 addr
    (5)  addr \in w_2 \or b \in a_2 ==> get \mu' addr = get \mu_2 addr
    (6)  addr ~: w_1\<union>w_2 & b ~: a_1\<union>a_2 ==> get \mu' addr = get \mu addr
  \end{lstlisting}
  The properties (1)--(3) define the state of blocks in the combined memory:
  a fresh block in \is{\mu'} was fresh already in \is{\mu}, and has not been allocated (1);
  an allocated block was already allocated or has been allocated, but has not been freed (2); and a freed block was already freed, or has been freed (3).
  The properties (4)--(6) define the content: addresses written or allocated in the first or second execution get their content from \is{\mu_1} (4) or \is{\mu_2} (5) respectively.
  Addresses not written nor allocated at all keep their original content (6).
